---
title: 'R Notebook: Project One'
author: "Mary E Rudis"
date: "1/21/2018"
output:
  html_document:
    df_print: paged
---

## Data 210G - Project One - DUE February 1, 2018

This is the Project One description using an R Notebook file. You can see the advantages of using this format. Instead of "ProjectOne.R" you may instead submit projects in either R Markdown or R Notebook format. The filename will be "ProjectOne.Rmd"

Run the following in RStudio:

```{r}
mpg1 <- mtcars[1]
mpg2 <- mtcars$mpg
variables <- names(mtcars)
vehicles <- rownames(mtcars)

```

Observe there are now several items listed under "Environment"

Type comments about each of these four objects. What kind of object is each thing?

For example, the following shows code **plus comment inserted:**

```{r}
  Data1 <- 1:20    ## R creates a vector of integers from 1 to 20 with length 20
```

Now include the following lines of code:

```{r}
is.vector(mpg1)
is.data.frame(mpg1)
is.vector(mpg2)
```

Run and comment on the result. How are mpg1 and mpg2 the same? How are they different?

For this part, you will need to insert code chunks into your .Rmd file as needed.

1) Store the data from the R dataset mtcars into an object named motor_trend_data

2) Calculate the smallest value, largest value, and average of mpg as well as hp.

3) Create a list of length 19 that starts with the 7 colors of the rainbow, ie red, orange, yellow, green, blue, indigo, violet, the numbers 1 through 10, and then TRUE, FALSE. Create it so that the list appears as a list of 3 in the global environment.

4) Type a line of code (a function) that returns (finds and records) the 5th item in the list.
